{In Example \ref{costrevenueprofitex1}, the price of a dOpi media player, in dollars per dOpi, is given as a function of the weekly sales $x$ according to the formula $p(x) = 450-15x$ for $0 \leq x \leq 30$.

\begin{enumerate}

\item  Find $p^{-1}(x)$ and state its domain.

\item  Find and interpret $p^{-1}(105)$.

\item  In Example \ref{costrevenueprofitex1}, we determined that the profit (in dollars) made from producing and selling $x$ dOpis per week is $P(x)= -15x^2+350x-2000$, for $0 \leq x \leq 30$.  Find $\left(P \circ p^{-1}\right)(x)$ and determine what price per dOpi would yield the maximum profit.  What is the maximum profit?  How many dOpis need to be produced and sold to achieve the maximum profit?


\end{enumerate}}
{\begin{enumerate}

\item $p^{-1}(x) = \frac{450-x}{15}$.  The domain of $p^{-1}$ is the range of $p$ which is $[0,450]$

\item  $p^{-1}(105) = 23$. This means that if the price is set to $\$105$ then $23$ dOpis will be sold.

\item $\left(P\circ p^{-1}\right)(x) = -\frac{1}{15} x^2 + \frac{110}{3} x - 5000$, $0 \leq x \leq 450$.  The graph of $y = \left(P\circ p^{-1}\right)(x)$ is a parabola opening downwards with vertex $\left(275, \frac{125}{3}\right) \approx (275, 41.67)$.  This means that the maximum profit is a whopping $\$41.67$ when the price per dOpi is set to $\$275$.  At this price, we can produce and sell $p^{-1}(275) = 11.\overline{6}$ dOpis.  Since we cannot sell part of a system, we need to adjust the price to sell either $11$ dOpis or $12$ dOpis. We find $p(11) = 285$ and $p(12) = 270$, which means we set the price per dOpi at either $\$285$ or $\$270$, respectively.  The profits at these prices are $\left(P\circ p^{-1}\right)(285) = 35$ and  $\left(P\circ p^{-1}\right)(270) = 40$, so it looks as if the maximum profit is $\$40$ and it is made by producing and selling $12$ dOpis a week at a price of $\$270$ per dOpi.

\end{enumerate}}